Research in the field of electricity consumption forecasting has established many methods for producing forecasts. These methods can roughly be divided into two categories; conventional methods and AI-based methods. Conventional methods such as time series analysis and regression methods have previously been widely applied to solve the problem of electricity consumption forecasting. Nowadays, AI-based methods are more popular due to their strength in solving nonlinear problems, something that the conventional methods struggle with \cite{mat_daut_building_2017}.

[Two comprehensive reviews on electricity consumption forecasting have been identified:

\begin{itemize}
    \item An extensive review and comparison of both statistical and ML/DL techniques for forecasting is found in \cite{deb_review_2017}. They also review combinations of different techniques, i.e. hybrid models. Claims that ANN has more advantages than statistical models, and has better performance for nonlinear problems. Highlights that hybrid models can be beneficial to capture complexities in building energy and operational data.
    \item Another review of statistical, AI, and hybrid methods for forecasting is found in \cite{mat_daut_building_2017}. They also highlight the strength in AI models for dealing with nonlinear patterns. Claims that hybrids between AI and Swarm Intelligence (SI) methods show potential for increased accuracy. Provides a clear overview of different studies regarding prediction time intervals, included features, building types etc.]
\end{itemize}

%\section{Conventional Approaches}
%probably not relevant since i wont be using it

\section{Machine Learning Approaches?}

[Forecasting daily cooling energy using ANN (for three university buildings): \cite{deb_forecasting_2016}]

\section{Monthly Forecasting}

[Could possibly use \cite{dong_applying_2005} as an ``early'' example of monthly forecasting. They use SVM, but do not handle seasonality in any certain way.]

Forecasting electricity consumption in the monthly scope requires handling complex non-linear patterns and seasonal tendencies \cite{cao_support_2016}. Monthly consumption is affected by factors such as economy, temperature, and season, which makes the consumption exhibit a mixture of trend, periodicity, and randomness \cite{tang_gm11_2022}. Additionally, when predicting the consumption of small-scale users, like in this thesis, the power demand is more sensitive to random factors. Human behavior may change periodically, both due to social factors such as holidays and summer vacations, and due to natural factors like the seasons \cite{sun_monthly_2019}. 

To address the seasonality of monthly forecasting, models incorporating seasonal indexing (SI) have been proposed. Cao and Wu \cite{cao_support_2016} aimed to improve monthly electricity consumption forecasts using support vector regression (SVR) in combination with SI and a novel evolutionary algorithm called fruit fly optimization (FOA). They calculated the seasonal index for each month, which was then used to adjust the predicted values according to the seasonal tendencies. FOA was applied to find the best values for parameters of the SVR model. Their results showed that both SI and FOA was able to effectively improve the forecasting accuracy, suggesting that the hybrid approach is a viable option for monthly forecasting.

Tang et al. \cite{tang_gm11_2022} proposed a model specifically designed to handle sparse data with periodic properties. Their model also incorporated SI to address the seasonal component. A GM(1,1) gray model, a mathematical optimization model for incomplete data, was used to predict the seasonal index. The SI was combined with a trend forecast value to obtain the final prediction, and results showed that the model significantly improved accuracy, even for small electricity consumption datasets.

Sun et al. \cite{sun_monthly_2019} approached the monthly forecasting problem by seasonal and trend decomposition of the electricity load time series, in combination with Auto-regressive Integrated Moving Average (ARIMA). This technique allowed for control of both the seasonal component's change rate and the trend component's smoothness, as well as increased robustness to outliers. Additionally, they incorporated economic factors into the forecast to improve accuracy. They found that the proposed model was able to increase forecasting accuracy, and determined that both the decomposition approach and inclusion of economic variables were effective to the forecast.

\section{Ensemble Learning Approaches}

Ensemble learning has previously been applied for the task of electricity consumption forecasting, mostly in the short-term scope. Divina et al. \cite{divina_stacking_2018} proposed a stacking ensemble learning approach for short-term  forecasting of Spain's general electricity consumption. They used a two-layer approach based on a bottom and a top layer. The bottom layer consisted of three base learning methods; regression trees based on evolutionary algorithms, Artificial Neural Network (ANN), and Random Forest (RF). The top layer used Generalized Boosted Regression Models (GBM) to combine the predictions of the base learners, and produce the final predictions. They found that the combination of weaker learning methods was able to produce superior results compared to several baselines, showing that ensemble learning is suitable for short-term electricity consumption forecasting. Neural networks were able to predict with high accuracy in the very near future, but accuracy declined when predicting further in the future. The ensemble method also showed significantly better performance compared to baselines when the amount of historical data was small, indicating that ensemble methods may be more flexible for different quantities of data.

Pinto et al. \cite{pinto_ensemble_2021} also utilized ensemble learning for forecasting electricity consumption in the short-term. Using consumption data from an office building, along with other variables such as temperature and humidity, they predicted hour-ahead consumption with three different ensemble methods. These methods were random forests, gradient boosted regression trees (GBR), and AdaBoost. They found that ensemble learning methods, especially AdaBoost, outperformed benchmark models such as support vector regression and fuzzy rule-based methods. They highlight including temperature and humidity information as advantageous to the forecasts, and they suggest that future work should consider additional external variables such as solar irradiation and thermal sensation.

Khan et al. \cite{khan_ensemble_2021} presented a stacking deep learning ensemble for short-term forecasts. The lowest level of the ensemble was compromised of multiple individual Long Short Term Memory (LSTM) models responsible for producing outputs. The top level of the ensemble was a Gated Recurrent Unit (GRU) responsible for increasing accuracy by combining outputs from the previous layer. They also utilized clustering to identify customer groups with similar profiles and load patterns. Based on hourly residential consumption data, they predicted hourly, daily, and weekly consumption, and found that the proposed model outperformed benchmark deep learning models. The clustering of customers helped with reducing model errors. Additionally, it was found that their model performed best for hour-ahead forecasts, while errors increased when approaching the weekly scales.

Hadjout et al. \cite{hadjout_electricity_2022} used an ensemble deep learning approach to forecast monthly electricity consumption for high voltage customers in the Algerian economic sector. They utilized three different deep learning models \hyp{} LSTM, GRU, and Temporal Convolutional Networks (TCN) \hyp{} and grid search to find optimal weights for each model in the final ensemble prediction. Compared to the individual models separately, the ensemble approach performed significantly better according to MAE and MAPE metrics.

Long-term forecasting using ensemble learning has been researched by Chen et al. \cite{chen_novel_2018}. They predicted the annual electricity consumption for American households using extreme gradient boosting forests and deep networks as base learners, and ridge regression as a combiner method. The data included variables pertaining to household demographics, household type, appliances, and geographical data, providing a wide range of information. The ensemble framework was able to reduce errors by 30\% compared to classical regression models such as linear regression and support vector machines. Investigation into the importance of input features using principal component analysis showed that electricity price, number of rooms, natural gas usage, and temperature during summer were among the variables that contributed the most information. They also note that some features, like number of rooms, number of bedrooms, and total square footage, have a strong correlation and have potential to be represented more efficiently. They argue that feature selection is necessary, since discarding of unimportant features can enhance model performance.

\section{Explainable AI Approaches}

Explainable AI methods have previously been applied for the task of short-term electricity consumption forecasting. Sim et al. \cite{sim_explainable_2022} proposed a novel XAI approach for variable selection in energy consumption forecasting. They analyzed the influence of 15 different input variables pertaining to time information, climate data, and past energy consumption data. LSTM was selected as prediction model and SHAP was used to interpret the variables. The findings showed that time data such as year and hour had a strong influence on the prediction results, along with temperature, surface temperature, and energy\hyp{}consumption\hyp{}difference. Variables with a weak or ambiguous influence included humidity, wind speed, and sunshine amount. They concluded that forecasting performance can be significantly improved by selecting high-impact variables based on XAI analysis.

Maarif et al. \cite{maarif_energy_2023} investigated the effect of 9 different features on energy consumption forecasting for a South Korean steel company. The features provided information about historical consumption, lagging and leading current power, carbon dioxide emissions, and load type. They utilized various LSTM models to make hour-ahead forecasts, and they used SHAP for the XAI analysis. It was found that leading current reactive power and seconds from midnight strongly influenced the model output, while CO$_2$ and load type had no effect on the forecast. They note that it is important to consider interpretability and performance trade-offs when developing XAI forecasting models, and that further research into this topic is needed for understanding the applicability of these models in the energy sector.

Januja et al. \cite{janjua_enhancing_2024} maybe, but not reputable journal.

Another hourly electricity consumption forecasting study was performed by Lazzari et al. \cite{lazzari_user_2022}. They provided a novel perspective by incorporating user behavior to enhance the accuracy of residential energy consumption forecasting. User behavior clusters were identified, and the day-ahead behavior was predicted using an XGBoost model. The predicted behavior, along with historical consumption, was then provided to an ANN model responsible for predicting the hourly day-ahead consumption. Compared to an ANN without the user behavior input, the proposed model was able to reduce MAPE and NRMSE by 7\% and 9\% respectively. A feature importance analysis using SHAP showed that calendar variables, namely week and season, had the greatest importance for the model. Temperature, notably minimum and maximum daily temperature, also had a large influence on model output. 

Moon et al. \cite{moon_advancing_2024} utilized ensemble learning and XAI methods for short-term residential building electricity consumption forecasting. They performed an extensive comparison between five ensemble learning models and ten deep learning models. It was found that the decision tree-based ensemble methods were superior in handling diverse and noisy data, rendering them useful for electricity consumption forecasting. They highlight robustness, interpretability, and efficiency as advantageous characteristics in ensemble models. They argue that ensemble models can handle data that is not extensively preprocessed and that they require less data than deep learning models. A feature importance and SHAP analysis was performed on several ensemble models, revealing that temperature-humidity index and wind chill temperature had significant importance for the forecasting. The SHAP analysis combined with the decision tree-based models provided interpretive insights about complex interactions in the data.

XAI approaches have also been applied to long-term forecasting \cite{machlev_explainable_2022}. Chakraborty et al.
\cite{chakraborty_scenario-based_2021} predicted long-term cooling energy consumption for residential and commercial buildings under different climate change scenarios. They used SHAP to explain the results and concluded that the XAI method increased the interpretability of the XGBoost prediction model. Another long-term study was conducted by Wenninger et al. \cite{wenninger_explainable_2022}, who predicted the annual energy consumption for German households, and proposed a novel QLattice model for enhanced explainability. They found that the QLattice model was appropriate for the task of energy consumption forecasting, even though XGBoost achieved slightly better predictive performance. However, a significant strength with the QLattice algorithm is its transparent design, which was found to greatly increase the model explainability compared to established models such as ANN, XGBoost, and Support Vector Regression (SVR).

    [With this, it is clear that there is a gap in research on XAI for medium-term forecasting. And in general, the consensus is that more research in the XAI field in the energy sector is needed (?)]